Staged Abstract Interpreters

### 2019-09-03 paper/80_related.tex
revert to as sign; typo

\paragraph{Optimizing Static Analysis Through Specialization}
The underlying idea in this paper is closely related to abstract
compilation (AC) \cite{Boucher:1996:ACN:647473.727587}: removing the
interpretative overhead on traversing the syntax tree by
specialization. Specifically, the residual program of AC can be either
text or closures. However, as we studied in Section \ref{cs_ac},
with the perspective of generative programming and monadic
interpreters, we make it more general and extensible, and easier.
\citet{Johnson:2013:OAA:2500365.2500604} adapt the idea of closure
generation to optimize small-step abstract interpreter in
state-transition style. The analyzed program is first compiled to an
IR called "abstract bytecode", which is actually a series of higher-order
functions which will be executed later on an abstract abstract machine
for that IR. \citet{damian1999partial} provides a formal treatment to
abstract compilation and Shiver's CFA, as well as proofs to establish
the correctness of the certified specialized analyzer.
\citet{amtoft1999partial} applied partial evaluation for
constraint-based control flow analysis.

Splitting an analysis into multiple stages has also been studied for
analysis other than abstract-interpretation-based control flow
analysis, though the formulation is quite
different. \citet{DBLP:conf/cgo/HardekopfL11} apply staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
program code to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses are implemented by
generating constraints firstly and then let a solver discharge these
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Similar to the idea in this paper, such analyses can be split to
staged.  One piece of future work is to explore whether these
analyses can be implemented and optimized from
meta-programming. Souffl{\'e}, one of the state-of-the-art Datalog
engines that is commonly used in points-to analysis
\cite{Antoniadis:2017:PDS:3088515.3088522}, uses the idea of Futamura
projections for efficient implementation \cite{10.1007/978-3-319-41540-6_23}.
However, it only specializes Datalog rules to the running engine, instead
of a full pipeline from program AST to generating and
solving constraints.

Recent work shows that the numerical abstract domains can be
specialized w.r.t. the static program structure, although not by
way of staging. For example, decomposing polyhedrons
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143} to
smaller ones that depend on the variables involved in a statement can
significantly improve the running time and space of abstract
transformers.

\paragraph{Construction of Abstract Interpreters} Abstract interpretation
was proposed as a semantics-based approach to build sound static analysis by
approximation \cite{DBLP:conf/popl/CousotC77}. As for building semantic
artifacts, the Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach
shows that abstract interpreters can be derived systematically from
concrete semantic artifacts. The AAM approach is closely related to
control-flow analysis for higher-order languages
\cite{Midtgaard:2012:CAF:2187671.2187672}.  Using monads to construct
abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17,
  Darais:2015:GTM:2814270.2814308}.  The unstaged abstract interpreter
in this paper follows the abstracting definitional interpreters
approach from \citet{DBLP:journals/pacmpl/DaraisLNH17}.  Similar to
definitional abstract interpreters,
the reconstruction of big-step abstract interpreters with delimited
continuations shown by \citet{Wei:2018:RAA:3243631.3236800},
\citet{Keidel:2018:CSP:3243631.3236767} present a modular, arrow-based
abstract interpreter that allows proof of soundness to be constructed
compositionally. \citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis. We borrow the notation of modular analysis for our approach.
However the staged abstract interpreter does not generate complete summaries for
modules; instead, it is able to specialize the analysis modularly. Similar, AC
is considered as an example of this kind of modular analysis \cite{DBLP:conf/cc/CousotC02}.

\paragraph{Control-flow Analysis} Our case studies and evaluation focus on
control-flow analysis for functional languages
\cite{Shivers:1991:SSC:115865.115884, Midtgaard:2012:CAF:2187671.2187672}. In
general, $k$-CFA (where $k > 0$) is considered impractical to use as a compiler
pass, due to its EXPTIME lower bound \cite{VanHorn:2008:DKC:1411204.1411243}.
However, several variants of 0CFA were developed and shown to be useful in
practice \cite{Adams:2011:FTR:2048066.2048105,
Bergstrom:2014:PEH:2628136.2628153, ashley:practical, Reppy:2006:TCA:1159876.1159888}.

\paragraph{Two-level Semantics} The idea of reinterpreting the semantics as
abstract interpretation can be traced to Nielson's two-level semantics
\cite{NIELSON1989117}; using two-level semantics for code generation
was also explored by \citet{NIELSON198859}.
\citet{Sergey:2013:MAI:2491956.2491979}'s work of monadic abstract
interpreters is also closely related to the two-level semantics: the
use of a generic interface with monads and then reinterpreting it with
different semantics is already two-level. Instead of focusing on
semantics, this paper shows how a staged analyzer can be built and
used to increase efficiency of static analysis. We augment the monadic
abstract interpreter from another dimension, using code generation to
produce efficient low-level code. Hence the presented work can be
considered as an two-dimensional application of the two-level semantics.

\paragraph{Partial Evaluation and Multi-stage Programming}
Partial evaluation as an automatic technique for program
specialization was studied comprehensively by
\citet{10.1007/3-540-61580-6_11, DBLP:books/daglib/0072559}.  In this
paper, we use mutli-stage programming as an approach to program
specialization. The Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} we used in the paper relies on
type-level stage annotations.  Other notable implementations of MSP
exist in the ML family, e.g., MetaML \cite{DBLP:conf/pepm/TahaS97} and
MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03,
  DBLP:conf/flops/Kiselyov14}.  The idea of staging an abstract
interpreter we presented in this paper is also applicable to other MSP
systems.  Multi-stage programming has been widely used to improve the
performance in many domains, such as optimizing compilers or
domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14,
  DBLP:conf/snapl/RompfBLSJAOSKDK15,
  DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
  DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
  DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17,
  Ofenbeck:2017:SGP:3136040.3136060}, data processing
\cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
  DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.

As a source of inspiration of this paper, the Futamura projections
reveal a hierarchy and close relationship between interpreters and
compilers, which was originally proposed by
\citeauthor{futamura1971partial} in the 1970s\cite{futamura1971partial},
and later reprinted \cite{Futamura1999}.
\citeauthor{Amin:2017:CTI:3177123.3158140} considers a tower of
concrete interpreters and how to collapse them by using MSP -- it
would be interesting to explore this idea for multiple layers of
abstract interpreters \cite{Cousot:2019:AAI:3302515.3290355,
  Giacobazzi:2015:APA:2676726.2676987}.  \citet{10.1007/11561347_18}
discussed a let-insertion, memoizing monads for code generation, using
Gaussian Elimination in MetaOCaml as example. In this work,
let-insertion is handled by the ANF transformation in LMS, and the
memoizing monad here just combines a reader monad and state monad
operating on staged data.
Similar to the idea in this paper, specializing monadic interpreters
was explored by \citet{danvy1991compiling, DBLP:conf/dsl/SheardBP99},
but is different from the performance motivation here: the generated code
is in monadic form.

>>>>

\paragraph{Optimizing Static Analysis Through Specialization} The underlying
idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretative overhead of
traversing the syntax tree by specialization. The residual programs of AC can
be either textual (using quasi-quotations) or closures (using higher-order
functions).  Our approach is more flexible between representations and requires
less engineering efforts to refactor the analyzer (Section \ref{cs_ac}).
\citet{Johnson:2013:OAA:2500365.2500604} adapted closure generation to optimize
small-step abstract interpreter in state-transition style. The program
is first compiled to an "abstract bytecode" IR. The IR consists of higher-order
functions, which will be executed later on an abstract abstract machine for
that IR.  \citet{damian1999partial} provided a formal treatment to abstract
compilation and control-flow analysis, as well as proofs to establish the
correctness of a certified specialized analyzer.  \citet{amtoft1999partial}
applied partial evaluation for constraint-based control-flow analysis.

Splitting an analysis into multiple stages has also been studied for other
forms of static analyses. 
\citet{DBLP:conf/cgo/HardekopfL11} applied staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
programs to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses are implemented by
generating constraints firstly and then use a solver to discharge the
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Such analyses can be considered having stage distinctions.  One piece of future
work is to explore whether they can be implemented and optimized from
meta-programming. 
Souffl{\'e}, one of the state-of-the-art Datalog engines commonly used in
points-to analysis \cite{Antoniadis:2017:PDS:3088515.3088522}, uses Futamura
projections for efficient implementation \cite{10.1007/978-3-319-41540-6_23}.
However, it only specializes Datalog rules to the running engine, instead of a
full pipeline from a program AST to a specialized solver.

Recent work shows that numerical abstract domains can be
specialized w.r.t. the static program structure, although not by
way of staging. For example, decomposing polyhedrons
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143} to
smaller ones depending on the variables involved in a statement can
significantly improve the running time and space consumption of abstract
transformers.

\paragraph{Construction of Abstract Interpreters} Abstract interpretation is a
semantics-based approach to build sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77, Cousot98-5}. 
As for building semantic artifacts, the Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach shows that
abstract interpreters can be derived systematically from concrete semantic
artifacts based on operational semantics. 
The AAM approach is closely related to control-flow analysis for higher-order languages
\cite{Midtgaard:2012:CAF:2187671.2187672, Shivers:1991:SSC:115865.115884}. 
Using monads to construct abstract interpreters was explored by
\citet{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17,
Darais:2015:GTM:2814270.2814308}.
The unstaged abstract interpreter in this paper follows the abstracting
definitional interpreters (ADI) approach from
\citet{DBLP:journals/pacmpl/DaraisLNH17}.  Similar to ADI, reconstructing
big-step abstract interpreters with delimited continuations was shown by
\citet{Wei:2018:RAA:3243631.3236800}; \citet{Keidel:2018:CSP:3243631.3236767}
presented a modular, arrow-based abstract interpreter that allows soundness
proof to be constructed compositionally. 
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis. We borrow the notation of modular analysis to our approach.
However, the staged abstract interpreter does not generate complete summaries for
modules; instead, it is able to specialize the analysis modularly. Similarly, AC
is considered as an example of this kind of modular analyses \cite{DBLP:conf/cc/CousotC02}.

\paragraph{Control-flow Analysis} Our case studies and evaluation focus on
control-flow analysis for functional languages
\cite{Shivers:1991:SSC:115865.115884, Midtgaard:2012:CAF:2187671.2187672}. In
general, $k$-CFA (where $k > 0$) is considered impractical to be used as a
compiler pass, due to its EXPTIME lower bound
\cite{VanHorn:2008:DKC:1411204.1411243}.  However, several variants of 0CFA
were developed and shown to be useful in practice
\cite{Adams:2011:FTR:2048066.2048105, Bergstrom:2014:PEH:2628136.2628153,
ashley:practical, Reppy:2006:TCA:1159876.1159888}.

\paragraph{Two-level Semantics} The idea of reinterpreting the semantics as
abstract interpretation can be traced to Nielson's two-level semantics
\cite{NIELSON1989117}; using two-level semantics for code generation
was also explored by \citet{NIELSON198859}.
\citet{Sergey:2013:MAI:2491956.2491979}'s monadic abstract
interpreters is closely related to the two-level semantics: the
use of a generic interface with monads and then reinterpreting it with
different semantics is already two-level. Instead of focusing on
semantics, this paper shows how a staged analyzer can be built and
used to increase efficiency of static analysis. We augment the monadic
abstract interpreter from another dimension, using code generation to
produce efficient low-level code. Hence the presented work can be
considered as a two-dimensional application of the two-level semantics.

\paragraph{Partial Evaluation and Multi-stage Programming}
Partial evaluation is an automatic technique for program
specialization and was studied comprehensively by
\citet{10.1007/3-540-61580-6_11, DBLP:books/daglib/0072559}. 
In this paper, we use mutli-stage programming to specialize programs.
The Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} we used in the paper relies on
type-level stage annotations.  Other notable implementations of MSP
exist in the ML family, e.g., MetaML \cite{DBLP:conf/pepm/TahaS97} and
MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03,
  DBLP:conf/flops/Kiselyov14}.  
The idea of staging an abstract interpreter we presented here 
is also applicable to other MSP systems.  
Multi-stage programming has been widely used to improve the
performance in many other domains, such as optimizing compilers or
domain-specific languages \cite{DBLP:conf/pldi/RompfSBLCO14,
  DBLP:conf/snapl/RompfBLSJAOSKDK15,
  DBLP:journals/tecs/SujeethBLRCOO14, DBLP:conf/gpce/SujeethGBLROO13,
  DBLP:journals/jfp/CaretteKS09}, numerical computation \cite{PGL-038,
  DBLP:conf/pepm/AktemurKKS13}, generic programming
\cite{DBLP:journals/pacmpl/Yallop17,
  Ofenbeck:2017:SGP:3136040.3136060}, data processing
\cite{DBLP:conf/oopsla/JonnalageddaCSRO14,
  DBLP:conf/popl/KiselyovBPS17}, query compilation in databases
\cite{DBLP:conf/osdi/EssertelTDBOR18, DBLP:conf/sigmod/TahboubER18},
etc.

As a source of inspiration of this paper, the Futamura projections reveal a
close relation between interpreters and compilers, which was
originally proposed by \citet{futamura1971partial}, and later reprinted \cite{Futamura1999}.
\citeauthor{Amin:2017:CTI:3177123.3158140} considered a tower of concrete
interpreters and showed how to collapse them by using MSP -- it would be interesting
to explore this idea for multiple abstract interpreters
\cite{Cousot:2019:AAI:3302515.3290355, Giacobazzi:2015:APA:2676726.2676987}.
\citet{10.1007/11561347_18} discussed a monad with let-insertion and
memoization for code generation, using Gaussian Elimination in MetaOCaml as
example. In this work, let-insertion is handled by the ANF transformation in
LMS, and the memoizing monad is just a combination of a reader monad and a state monad
operating on staged data.  Similar to the idea in this paper, specializing
monadic interpreters was explored by \citet{danvy1991compiling,
DBLP:conf/dsl/SheardBP99}. But they are different from the performance motivation
here: the generated code is in monadic form.

### 2019-09-03 paper/90_conc.tex
misc update
revert to as sign; typo

In this paper, we present a simple and elegant framework that extends the first
Futamura projection to abstract interpreters.
To realize it, we design a generic interpreter that abstracts over both
binding-time and value domains; by instantiating it with staged monads for
abstract interpretation, the staged abstract interpreter can be readily derived.
The meta-programming approach brings new insight and implementation techniques
to optimize static analyzers. The optimization can be realized in a completely
orthogonal way to the interpreter and abstract semantics.
In our empirical evaluation, staging improves an order of magnitude of
performance in terms of running time.

>>>>

In this paper, we present a simple and elegant framework that extends the first
Futamura projection to abstract interpreters.
To realize it, we design a generic interpreter that abstracts over both
binding-times and value domains; by instantiating it with staged monads for
abstract interpretation, the staged abstract interpreter can be readily derived.
The meta-programming approach brings new insights and implementation techniques
to optimize static analyzers. Optimizations can be achieved orthogonally to 
the interpreter and abstract semantics.  In our empirical evaluation, staging
improves performance by an order of magnitude in terms of running time.

### 2019-09-03 paper/abstract.tex
revert to as sign; typo

It is well known that a staged interpreter is a compiler: specializing an
interpreter to a given program produces an equivalent executable that runs faster.
This connection is known as the first Futamura projection.
It is even more widely known that an abstract interpreter is a program analyzer:
tweaking an interpreter to run on an abstract domain produces a sound static
analysis. What happens when we combine these two ideas, and apply specialization to
an \emph{abstract} interpreter?

>>>>

It is well known that a staged interpreter is a compiler: specializing an
interpreter to a given program produces an equivalent executable that runs faster.
This connection is known as the first Futamura projection.
It is even more widely known that an abstract interpreter is a program analyzer:
tweaking an interpreter to run on abstract domains produces a sound static
analysis. What happens when we combine these two ideas, and apply specialization to
an \emph{abstract} interpreter?
